Trends in Mortality Rates

### 2024-09-25 paper/paper.qmd
new info added

Additionally, the challenging living conditions faced by homeless individuals, compared to those of wealthier individuals who also experienced hardships during this time, suggest that the mortality rates among shelter residents are likely to remain elevated. COVID-19 was also a primary driver behind the large number of deaths, as evidenced by the immediate decrease in the number of deaths following the post-pandemic period shown in the graphs. The data was examined more in depth by analyzing the mortality rate and average mortality rate between male and female homeless individuals in shelter residences. There are a lot more male death compare to female death, but this make sense as the proportion of male residents is greater than that of female residents, with 60% male and only 40% female (@toronto_shelter_flow_data). Additionally, the average age of deceased individuals is around 50 to 60 years old (@datacommons_life_expectancy). According to data, the life expectancy in Canada is approximately 81 years, indicating that the average age at death for homeless individuals is quite young.

@@

The data used in this paper is access in from Open Data Toronto and the particular data set used was the Deaths of Shelter Residents (@toronto_deaths_shelter_residents). To analysis the data and creating graphs using the data, following package that was build in the R program @citeR was used: tidyverse (@tidyverse), dplyr (@dplyr), knitr (@knitr), and ggplot2 ()@ggplot2.

@@

The Deaths of Shelter Residents data is part of a larger database shared by Toronto Public Health (TPH), which also includes deaths of homeless individuals living in shelters and outdoors. Therefore, the findings in this paper focus solely on homeless individuals residing in shelters, which may limit the dataset and lead to potential biases that do not fully represent the mortality of all homeless individuals.

@@

@fig-TotalNumberofDeathByYear is showing the total amount of decedent of homeless living in the Shelter.

@@

@fig-TotalDecedentsbyMonth shows the number of deaths in each month-year pair from 2007 to 2024. 
```{r}
#| label: fig-TotalDecedentsbyMonth
#| fig-cap: Yearly Mortality rate from 2007-2024
#| echo: false

@@

Talk way more about it.

@fig-TotalDecedentbyMonthcovid is a close up look from @fig-TotalDecedentsbyMonth focusing on data ranging from year 2019 to 2022, during the pandemic. 
```{r}
#| label: fig-TotalDecedentbyMonthcovid
#| fig-cap: Yearly Mortality rate from 2019-2022
#| echo: false

# Filter the data for years 2019 to 2022
Death_data_filtered <- Death_data %>%
  filter(year >= 2019 & year <= 2022)

@@

# Plot the graph for filtered data
Death_data_filtered |>
  ggplot(aes(x = YearMonth, y = total_decedents, group = 1)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  scale_x_discrete(breaks = Death_data_filtered$YearMonth[seq(1, nrow(Death_data_filtered), by = 6)]) + # Show every 6th month
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) + # Display x-axis labels vertically
  labs(title = "Trends in Total Decedents by Month and Year (2019-2022)",
       x = "Month-Year",
       y = "Total Decedents") +
  theme_minimal()
```

@ fig-TotalDecedentsFemaleVsMale display a comparison in number of death between females and males.
```{r}
#| label: fig-TotalDecedentsFemaleVsMale 
#| fig-cap: Total Decedents Female VS Male
#| echo: false
#| include: false
#| warning: false
#| message: false

@@

@fig-AvgAgeofDecedentsFemaleVSMale presents a comparison of the average age at death between females and males.
```{r}
#| label: fig-AvgAgeofDecedentsFemaleVSMale 
#| fig-cap: Total Decedents Female VS Male
#| echo: false
#| warning: false
#| message: false

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

@@

@@

@@

Weaknesses and next steps should also be included.

>>>>

The data was examined more in depth by analyzing the mortality rate and average mortality rate between male and female homeless individuals in shelter residences. There are a lot more male death compare to female death every single year. Additionally, the average age at death is around 50 to 60 years old, which is relatively young compared to Canada’s life expectancy. Future details about the findings will be discussed in the discussion section.

@@

The data used in this paper is access in from Open Data Toronto and the particular data set used was the Deaths of Shelter Residents (@toronto_deaths_shelter_residents). To analysis the data and creating graphs using the data, following package that was build in the (R program @citeR) was used: tidyverse (@tidyverse), dplyr (@dplyr), knitr (@knitr), and ggplot2 (@ggplot2).

@@

The Deaths of Shelter Residents data is part of a larger database shared by Toronto Public Health (TPH), which also includes deaths of homeless individuals living in shelters and outdoors (@toronto_homeless_deaths). Therefore, the findings in this paper focus solely on homeless individuals residing in shelters, which may limit the data set and lead to potential biases that do not fully represent the mortality of all homeless individuals.

@@

@@

```{r}
#| label: fig-TotalDecedentsbyMonth
#| fig-cap: Yearly Mortality rate from 2007-2024
#| echo: false

@@

```{r}
#| label: fig-TotalDecedentbyMonthcovid
#| fig-cap: Yearly Mortality rate from 2019-2023
#| echo: false

# Filter the data for years 2019 to 2022
Death_data_filtered <- Death_data %>%
  filter(year >= 2019 & year <= 2023)

@@

# Plot the graph for filtered data
Death_data_filtered |>
  ggplot(aes(x = YearMonth, y = total_decedents, group = 1)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  scale_x_discrete(breaks = Death_data_filtered$YearMonth[seq(1, nrow(Death_data_filtered), by = 6)]) + # Show every 6th month
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) + # Display x-axis labels vertically
  labs(title = "Trends in Total Decedents by Month and Year (2019-2023)",
       x = "Month-Year",
       y = "Total Decedents") +
  theme_minimal()
```

```{r}
#| label: fig-T 
#| fig-cap: Total Decedents Female VS Male
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: fig-AvgAgeofDecedentsFemaleVSMale 
#| fig-cap: Total Decedents Female VS Male
#| echo: false
#| warning: false
#| message: false

@@

## First discussion point {#sec-first-point}
@fig-TotalNumberofDeathByYear shows the total number of decedents among homeless individuals living in shelters. The graph visualizes two notable increases in 2013 and 2019. According to making connection with other sources, the primary causes of these increases are linked to macro health conditions occurring during those times.

In 2013, there was an influenza outbreak across Ontario that lasted from 2013 to 2016. Influenza is a viral infection that affects the nose, throat, and sometimes the lungs, leading to respiratory illness. Symptoms can range from mild to severe, and in some cases, the flu can result in death. The best way to prevent infection is to get vaccinated each year (@cdc_flu_key_facts). There is a clear trend of increasing deaths up until 2015, followed by a drop in 2016. This pattern closely resembles the total number of influenza outbreaks reported in the article "Influenza Outbreaks in Ontario Hospitals" (@cameron2019). According to Table 1 in the article, the number of outbreaks increased from 38 in 2013-2014 to 117 in 2014-2015, then decreased to 36 in 2015-2016. This rise and fall in outbreaks corresponds with the trends observed in @fig-TotalNumberofDeathByYear, where the number of deaths peaked in 2015 and then significantly declined in 2016, coinciding with the flu being brought under control.

One notable point is that in the article, the fatality rate for the 2014-2015 outbreak was only 2.4%, despite the hundreds of flu outbreaks that occurred that year. However,@fig-TotalNumberofDeathByYear shows that the number of deaths in 2015 was around 45. This discrepancy can be explained by the fact that the data in the article pertains to hospital outbreaks, where individuals have access to better medical care. In contrast, homeless individuals may lack the resources to seek treatment, leading to a higher number of deaths in this population.

@@

Another clear increasing trend in the mortality rate is evident in 2019, as shown in @fig-TotalNumberofDeathByYear. This is not surprising, given that people around the world were experiencing and still recovering from the COVID-19 pandemic, which began in 2019 and continued until the end of 2023. To better understand the impact, the Government of Canada (@canada_health_infobase) continuously updates the numbers of cases and deaths related to the pandemic, showing a significant rise in fatalities from 2021 to 2022, followed by a decline after 2023.

Additionally, the challenging living conditions faced by homeless individuals, compared to those of wealthier individuals who also experienced hardships during this time, suggest that the mortality rates among shelter residents are likely to remain elevated. COVID-19 was also a primary driver behind the large number of deaths, as evidenced by the immediate decrease in the number of deaths following the post-pandemic period shown in @fig-TotalDecedentbyMonthcovid.

@@

From @fig-T, it is evident that the mortality rate for males is significantly higher than that for females. One reason for this observation is that the proportion of male residents in shelters is greater than that of female residents, with 60% male and only 40% female (@toronto_shelter_flow_data). Nonetheless, research has shown that females tend to live longer than males at all stages of life, from infancy to young adulthood and into retirement (@ourworldindata_life_expectancy). The overall reason for the higher male mortality rate compared to females can be attributed to genetic factors as well as differences in health awareness and behaviors between the two genders.

@fig-AvgAgeofDecedentsFemaleVSMale indicates that the average age at death for homeless individuals living in shelters ranges from 50 to 60 years. According to data from the World Bank, the life expectancy in Canada is approximately 81 years, which suggests that the average age at death for homeless individuals is relatively young (@datacommons_life_expectancy). Despite the increasing life expectancy in today’s society, as health and technology continue to evolve, homeless individuals still die at a younger age due to financial constraints and limited access to resources.

@@

The findings in this paper represent a small aspect of what is happening in the world, focusing specifically on the homeless population. While these results cannot be used to draw definitive conclusions about broader trends, they do provide insights into patterns that may be occurring on a larger scale.

The connections discussed in this section are not statistically proven, as no models were run to demonstrate the correlation between the deaths of homeless individuals residing in shelters and the various possible explanations mentioned. In addition, the actual causes of death may remain unknown due to confidentiality restrictions and are inferred based on global events that could have contributed to these fatalities.

Other data will have to be gathered and analyzed using appropriate models to establish a solid scientific connection.
For instance, to establish a connection between the increase in death numbers in 2013 and 2019 and the flu outbreak and pandemic, data on flu-related deaths and pandemic-related deaths should be gathered and analyzed. A model should be run to test the relationship between these variables.
